\chapter*{Abstract}

The Nintendo Wii is well-known for its innovative, motion-based controls and engaging, family-friendly games such as Mario Kart Wii. Despite its hardware limitations compared to modern consoles, its local multiplayer experiences have cultivated a devoted following. However, with the rapid shift toward online gaming, recreating the Wii’s in-person, split-screen experiences has become increasingly challenging. This project proposes a solution that vitalises the Wii’s input and output interfaces, enabling remote players to enjoy an experience that mirrors local multiplayer gaming.

The approach centres on two key components. First, video and audio streaming techniques capture the Wii’s outputs and deliver them to remote devices using low-latency protocols. This ensures fluid gameplay and preserves the authenticity of the original experience. Second, a novel controller input relay system transmits Wiimote signals, including motion and button inputs, over a network. This system addresses challenges such as Bluetooth communication, network variability, and precise synchronisation between audiovisual and control data, ensuring real-time responsiveness.

By bridging the gap between traditional local multiplayer and modern online connectivity, this project extends the life of a beloved console while revitalising classic gaming experiences. Furthermore, it establishes a framework for adapting retro systems to contemporary, distributed gaming environments. The work not only preserves the social and communal essence of local play but also offers broader implications for making nostalgic gaming experiences accessible to players across geographically separated locations.
