\chapter*{Declaration}
I declare that the material submitted for assessment is my own work except
where credit is explicitly given to others by citation or acknowledgement. This
work was performed during the current academic year except where otherwise
stated.

%% TODO: Add word count
The main text of this project report is [TODO: Add word count] words long

In submitting this project report to the University of St Andrews, I give
permission for it to be made available for use in accordance with the
regulations of the University Library. I also give permission for the report to
be made available on the Web, for this work to be used in research within the
University of St Andrews, and for any software to be released on an open source
basis as long as proper credit is given to the authors.

%TODO: Check if proper credit part and ownership of resulting IP is correct

I retain the copyright in this work, and ownership of any resulting intellectual property.

\RedeclareSectionCommand[beforeskip=1cm,afterskip=1cm]{chapter}
{\let\clearpage\relax \chapter*{Acknowledgements}}
I would like to thank my supervisor, Dr. Tom Spink, for his encouragement, support, and counsel throughout my Senior Honours project. I would also like to thank my girlfriend, Angelika Blaszczak, for her endless love, patience, and understanding during both the challenging and triumphant moments of this journey. In addition, I am deeply grateful to my family and friends for their constant support and encouragement throughout my time at the University of St Andrews. Finally, I would like to thank the School of Computer Science for providing a stimulating academic environment, excellent resources, and insightful feedback that were essential to the successful completion of this project and my undergraduate studies.
\RedeclareSectionCommand[beforeskip=0pt,afterskip=1cm]{chapter}
