\chapter{Plan and Objectives from the Semester 2 Mid-Semester Report}
\label{appendix:plan-d1-semester2}

\section{Plan}
On a high level, I plan to continue developing the rewriter/the “rewriting
engine”. This will involve implementing the clean/dirty flag and using it to
optimise the rewriting algorithm, this will require metadata within all
Expression nodes. Additionally, I plan to add warning messages, tracing and a
debug mode to the rewriter as described in \hyperfootnote[this
issue]{https://}{www.github.com/conjure-cp/conjure-oxide/issues/181}. There are
also many more optimisations that can be made to the rewriter past the simple
clean/dirty flag optimisation. I plan to research more optimisation techniques,
particularly those found within Savile Row, and implement them for the
rewriter.

It will also be necessary to work on features of the rewriter that are
requested by other members of the rust team. For example, there may be a need
to look at the parents of Expression nodes in order to properly apply rules
that need extra context. I plan to work on these features as they are requested
by the rest of the Rust team. I also plan to continue working on the
documentation for the rewriter and the rest of the Rust code base.
\subsection{Primary objectives}
\begin{itemize}
    
    \item Implement metadata and add it to all Expression variants
    \item Implement the clean/dirty flag and use it to optimise the
        rewriting algorithm
    \item Add warning messages, tracing and a debug mode to the rewriter as per
        \hyperfootnote[this
        issue]{https://}{www.github.com/conjure-cp/conjure-oxide/issues/181} 
\end{itemize}
\subsection{Secondary objectives}
\begin{itemize}
    
    \item Research optimisation techniques, particularly those found
        within Savile Row, and implement them for the rewriter
\item Work on features of the rewriter that are requested by the rest of
    the Rust team
\item Continue working on the documentation for the rewriter and the rest
    of the Rust code base
\end{itemize}
